---
title: "Data products final presentation"
author: "Szymon Tomczyk"
date: "10/23/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r include=FALSE}
library(tidyverse)
library(GGally)
```

## The "diamonds" R dataset
This dataset contains the the prices and other attributes of almost 54,000 diamonds.
```{r message=FALSE, warning=FALSE, comment=""}
data(diamonds)
head(diamonds)
```
My shiny app will use this dataset to visualise the relationship between various properties of the diamond and it's price as well as build a smimple linear model invosetigating the relationship between the price and other variables

## A closer look at diamonds dataset and the variables used in the shiny app
```{r include=FALSE}

diamonds %>% select(price, carat:clarity) -> diamonds
```
```{r message=FALSE, warning=FALSE}
ggpairs(diamonds)
```

## My shiny app looks like this
```{r}
knitr::include_url("https://sztomczyk86.shinyapps.io/data_products_final_v2_ST/")

```

## Thank you for cheking out my presentation