---
output:
  pdf_document: default
  html_document: default
---
# Statistical Inference Course Project Part 2
_by Szymon Tomczyk_

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(datasets)
```

## Part 2 Basic Inferential Data Analysis
_In this part we will use the ToothGrowth dataset studying the impact of dose and delivery method of Vitamin C on tooth growth in guiney pig. I will perform an exploratory analysis, look at the summary of the data and do some basic hyptotheis testing._

### Loading the data
```{r echo=TRUE, message=FALSE, warning=FALSE}
data("ToothGrowth")
```

### Exploratory analysis
Look at the structure of the data and the summary statistics
```{r}
str(ToothGrowth)
summary(ToothGrowth)
```
Plot of the Lenght of the odontoblasts by the method of delivery: VC - ascorbic acid, OJ - orange juice. 
```{r}
ggplot(ToothGrowth, aes(x = as.factor(dose), y = len)) +
        geom_boxplot(aes(fill=supp)) +
        scale_fill_manual(values=c("#56B4E9", "#E69F00"))+
        facet_wrap(.~supp)+
        labs(title = "Lenght of odontoblasts by the dose of Vitamic C") +
        xlab("Dose (mg/day)") +
        ylab("Length of odontoblasts") +
        theme(legend.position = "none") 
```

### Hypothesis Testing

Test if there is a difference in tooth growth between the group that was given ascorbic acid and orange juice.
```{r}
t.test(ToothGrowth$len ~ ToothGrowth$supp)
```
_The 95% confidence interval contains the difference between the means of two groups so we cannot reject the null hipothesis_
 
Test if there is difference in tooth growth between the group that was given low (0.5) and high (2) dose of Vitamin C.
```{r}
ToothGrowth.sub <- filter(ToothGrowth, dose  %in% c(.5, 2))
t.test(ToothGrowth.sub$len ~ ToothGrowth.sub$dose)
```
_The observed p-value is very low, much lower than alpha = 0.05. Based on this fact we cannot assume that the means of the populations are equal. We reject the null hypthesis_

### Conclusions

Given the data we analyzed we can conclude that:

1. The method of delivery does not impact the tooth growth in guinea pig
2. The dose impacts the tooth growth in guinea pig

### Assumptions

1. The samples come from independent identicaly distributed populations
2. The samples are representative of the population
