---
title: "Reproducible research Assignment 1 by S. Tomczyk"
output: pdf_document
---
## Load neccessary packages
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(Hmisc)
library(gridExtra)
```



## Downloading and loading data into R

The script will check if the input file exist in your workingin directory and if 
it does not it will download it. Then the file will be unzipped and loaded into R

```{r}
if(!file.exists("activity.csv")) {
        
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
              "data.zip")
        
        unzip("data.zip")

         activity <- read.csv("activity.csv", header = TRUE, 
                             colClasses = c("numeric", "Date", "character"))

} else {
        
        activity <- read.csv("activity.csv", header = TRUE, 
                             colClasses = c("numeric", "Date", "character"))
        
}
```

### Look at the structure and summary of the loaded data
```{r}
str(activity)
summary(activity)
```

### Calculate the sum of the steps taken each day

```{r}

stepsDaily <- activity %>% group_by(date) %>% summarise(sum = sum(steps))

head(stepsDaily)

```

Produce histogram of the total number of steps taken per day
```{r warning=FALSE}

ggplot(data = stepsDaily, aes(sum)) + 
        geom_histogram(binwidth = range(stepsDaily$sum, na.rm = TRUE)[2]/30) +
        xlab("Total number of steps") + ylab("Number of Days")+
        labs(title="Hisogram of the total number of steps per day")
```

### Calculate mean and median number od steps taken each day.
```{r}
stepsM  <-activity %>% group_by(date) %>% 
        summarise(mean = mean(steps, na.rm = TRUE), 
                  median = median(steps, na.rm = TRUE))

print(stepsM)
```

### Plot the avergage number of steps taken each day
```{r warning=FALSE}

ggplot(data = stepsM, aes(date, mean))+
        geom_point() + geom_smooth() +
        labs(title = "Average number of steps per day") +
        xlab("Day") +
        ylab("Average number of steps")

```

### Find the 5 minute interval that on average had maximum number of steps

```{r}
stepsMax <- activity %>% group_by(interval) %>% 
        summarise(mean = mean(steps, na.rm = T)) %>%
        filter(mean==max(mean)) %>% print()
```

The 5 minute interval with highest average number of steps was 8.35 in the morning

### Impute the missing data based used random values from the range of recorded steps
```{r}
set.seed(666)
activity$imputed_steps <- impute(activity$steps, 'random')

head(activity)


```
  
Calculate total number of steps per day for the imputed data

```{r}
stepsDaily.imp <- activity %>% group_by(date) %>% summarise(sum = sum(imputed_steps))

head(stepsDaily.imp)
```

Compare the histograms of total number of steps with or without data imputation

```{r message=FALSE, warning=FALSE}
hist <- ggplot(data = stepsDaily, aes(sum)) + 
        geom_histogram(binwidth = range(stepsDaily$sum, na.rm = TRUE)[2]/30) +
        xlab("Total number of steps") + ylab("Number of Days")+
        labs(title="Hisogram of the total number of steps per day")+
        ylim(0, 10)
        

hist.imp <- ggplot(data = stepsDaily.imp, aes(sum)) + 
        geom_histogram(binwidth = range(stepsDaily$sum, na.rm = TRUE)[2]/30) +
        xlab("Total number of steps") + ylab("Number of Days")+
        labs(title="Hisogram of the total number of steps per day with imputed data") +
        ylim(0, 10)

grid.arrange(hist, hist.imp, nrow=2)
```

### Plot the average number of steps taken per 5-minute interval across weekdays and weekends

Add the WeekEnd WeekDay labels to the days using lubridate package
```{r}
activity$weekday <- wday(activity$date, label = T, abbr = F)

stepsWD <- activity %>% mutate(wd = if_else(weekday %in% 
                                c("Samstag", "Sonntag"), "WeekEnd", "WeekDay")) %>%
                        group_by(interval, wd) %>% 
                        summarise(mean = mean(steps, na.rm = TRUE))
stepsWD$interval <- as.numeric(stepsWD$interval)


```
Produce the plot showing the difference in number of steps between weekend and weekdays.

```{r}
ggplot(stepsWD, aes(interval, mean)) +
        geom_point(aes(colour = wd)) +
        ylab("Average number of steps")+
        xlab("5-minutes intervals")+
        labs(title = "Comparison between number of steps in Weekend vs Weekday")+
        facet_wrap(~wd, ncol=2)

```





